{\bf Landslide: A New Race-Finding Tool for 15-410}

### 2015-02-06 landslide-lecture/2015/lecture.tex
Add rough draft of landslide lecture for p2 for s15.

\title[Landslide]{{\bf Systematic Testing: A New Perspective on Races}}
\subtitle[]{ {\em more clever than } ``\texttt{slaughter cho2}'' {\em since 2011.}}
\author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

@@

% Say: Since I'm no longer a member of course staff, I'm allowed to give away a
% race condition that some of you might have, for the sake of example.

\begin{frame}[fragile]{Case Study}
	\begin{center}
	\begin{verbatim}
	    int thread_fork()
	    {
	        thread_t *child = construct_new_thread();
	        add_to_runqueue(child);
	        return child->tid;
	    }
	\end{verbatim}
	\end{center}

%	\begin{itemize}
%		\item On vanish, child's TCB is freed
%		\item Forking thread does use-after-free
%		\item Might return garbage instead of TID
%	\end{itemize}
\end{frame}

\definecolor{thread1}{RGB}{87,172,255}
\definecolor{thread2}{RGB}{255,201,102}

\begin{frame}{Thread Interleavings (``good'' case)}
	\begin{tabular}{|l|l|l}
		\cline{1-2}
		\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \\
		\cline{1-2}
		\texttt{spawn\_new\_thread} && \\
		\cline{1-2}
		\texttt{add\_to\_runqueue} && (new thread) \\
		\cline{1-2}
		\texttt{return child->tid} &&  \\
		\cline{1-2}
		& \texttt{vanish} & \\
		\cline{1-2}
		& (TCB gets freed) & (voluntary reschedule) \\
		\cline{1-2}
	\end{tabular}
\end{frame}

\begin{frame}{Thread Interleavings (race condition)}
	\begin{tabular}{|l|l|l}
		\cline{1-2}
		\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \\
		\cline{1-2}
		\texttt{spawn\_new\_thread} && \\
		\cline{1-2}
		\texttt{add\_to\_runqueue} && (new thread + preempted) \\
		\cline{1-2}
		& \texttt{vanish} & \\
		\cline{1-2}
		& (TCB gets freed) & (voluntary reschedule) \\
		\cline{1-2}
		\texttt{return child->tid} && (bad!) \\
		\cline{1-2}
	\end{tabular}
\end{frame}

@@

\begin{frame}{Testing Mechanisms}
	\textbf{Stress testing}: \texttt{slaughter cho2} and friends
	\begin{itemize}
		\item Attempting to exercise as many interleavings as practical
		\item Exposes race conditions at random
		\begin{itemize}
			\item ``If a preemption occurs at just the right time\ldots''
		\end{itemize}
		\item Cryptic panic messages or machine reboots
	\end{itemize}
	\linegap
	What if\ldots
	\begin{itemize}
		\item Make educated guesses about when to preempt
		\item Preempt enough times to run {\em every single} interleaving
		\item Tell the story of what {\em actually happened}.
		\item Overlook fewer bugs!
	\end{itemize}
	% Say: At the end of this lecture, I'm going to be looking for some among you...
\end{frame}

@@

	\linegap
	When do we fear being preempted?
	\begin{itemize}
		\item Threads becoming runnable (\texttt{fork()}, \texttt{cond\_signal()}, etc.)
			\begin{itemize}
				\item Preemptions may cause it to run before we're ready
			\end{itemize}
		\item Synchronization primitives (\texttt{mutex\_lock()}/\texttt{unlock()}, etc.)
			\begin{itemize}
				\item If buggy or used improperly\ldots
			\end{itemize}
		\item Unprotected shared memory accesses (``data races'')% Say: more on this later.
			\begin{itemize}
				\item May result in data structure corruption
			\end{itemize}
	\end{itemize}

@@

	{\bf About Landslide}
	\begin{itemize}
		\item Simics module, which traces:
			\begin{itemize}
				\item Every instruction executed
				\item Every memory access read/written
			\end{itemize}
			% FIXME for next year -- how much slowdown is there? 20x?
		\item Originally supported only P3s; can now test P2s as well
		\item {\em Landslide} shows that your {\em Pebbles} are less stable than you thought.
	\end{itemize}
\end{frame}

@@

\begin{frame}{Anatomy}
	\begin{center}
	\includegraphics[width=0.6\textwidth]{landslide.png}
	\end{center}
\end{frame}

\begin{frame}{Identifying Bugs}
	Landslide can {\em definitely discover}:
	\begin{itemize}
		\item Kernel panics % Say: note: the better your asserts are..!
		\item Deadlock
		\item Use-after-free / double-free
	\end{itemize}
	\linegap
	Landslide can {\em reasonably suspect}:
	\begin{itemize}
		\item Infinite loop (halting problem)
		\item Data races
	\end{itemize}
\end{frame}

\begin{frame}{Brief Aside: Data Races}
	A {\bf data race} is a pair of memory accesses between two threads, where:
	\begin{itemize}
		\item At least one of the accesses is a write
		\item The threads are not holding the same mutex
		\item The threads can be reordered (e.g. no \texttt{cond\_signal()} in between)
	\end{itemize}
	\pause
	\linegap

	Data races are {\em not necessarily} bugs, just highly suspicious!
	\begin{itemize}
		\item ``If threads interleaved the wrong way here, it might crash later.''
		\begin{itemize}
			\item Hmmm...
		\end{itemize}
	\end{itemize}
\end{frame}

\breakslide{
	\begin{center}
		\includegraphics[width=0.85\textwidth]{trace1.png}
	\end{center}
}
\breakslide{
	\begin{center}
		\includegraphics[width=\textwidth]{trace2.png}
	\end{center}
}

\begin{frame}{State Space Explosion}
	Serious problem: State spaces grow exponentially
	\begin{itemize}
		\item With $p$ preemption points and $k$ runnable threads, size $p^k$.
		\item Threatens our ability to explore everything.
		\item Fortunately, some sequences result in identical states.
	\end{itemize}
	\linegap

@@

\begin{frame}{Choosing the Right Preemption Points}
	So far we only {\em mitigated}, not {\em solved}, exponential explosion.
	\pause
	\linegap

@@

\begin{frame}{``Iterative Deepening'' (my current project)}
	Goal: Run the best tests for a given CPU budget.
	\linegap

	Based on experience from past 15-410 student volunteers
	\begin{itemize}
		\item Students worked best with an iterative process
		\item ``Start small, then add more preemption points as time allows''
		\item Can we automate this process?
	\end{itemize}
	\linegap

@@

@@

\section{Evaluation}
\breakslide{\Large Evaluation}

@@

\begin{frame}{Recruitment}
	\textbf{User study}: Come try Landslide on your P2!
	\begin{itemize}
		\item Bare minimum effort: No more than 30 minutes
		\begin{itemize}
			\item Clone a github URL, run setup script, run, send results
		\end{itemize}
		\item Testing P2s is fully automated, but can also test P3s
		\begin{itemize}
			\item Landslide requires manual kernel annotations
			\item Comes with pre-annotated reference kernel for P2
			\item Past students take approx. 1 hour to annotate own P3s.
		\end{itemize}
	\end{itemize}
	\pause
	\linegap

	{\bf Ultimate goal}: Provide Landslide to every 410 student during P2.
	\begin{itemize}
		\item This semester: Prep \& testing
			\begin{itemize}
				\item Improving user interface
				\item Making sure Landslide itself is stable
				\item Data collected won't be published
			\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame}{Recruitment}
	%User study this week, by appointment (\texttt{bblum@cs.cmu.edu})
	User study during exam week, by appointment (\texttt{bblum@cs.cmu.edu})
	\begin{itemize}
		\item (or anytime, really)
	\end{itemize}
	\linegap

%	Recommended to use fully-automated P2 test suite, but you can try testing your P3 with approx. 1 hour of manual annotation.
%	\linegap
%
	%Expect to spend:
	%\begin{itemize}
	%	\item Up to 3 hours, just to try it out.
	%	\item 4-6 hours, if you find a bug and track it down.
	%	\item More, for multiple bugs or the truly dedicated\ldots
	%\end{itemize}
	%\linegap

	Give feedback (intuitive? frustrating? found bugs?)
	\linegap

	{\bf \large Just for fun -- this will NOT impact your grade.}

	\linegap
	Watch your email for details!
\end{frame}

\section{Bonus Slide(s)}

\begin{frame}{More Results}
	\textbf{Is dynamically adding ``data race'' PPs effective?}
	\begin{itemize}
		\item 3 among reported bugs (9-15\%) required data races to expose.
	\end{itemize}
	\linegap

>>>>

\title[Landslide]{{\bf Landslide: A New Race-Finding Tool for 15-410}}
\subtitle[]{ {\em more clever than } ``\texttt{mandelbrot}'' {\em since 2011.}}
\author[Ben Blum]{Ben Blum \texttt{(bblum@andrew.cmu.edu)}}

@@

\definecolor{thread1}{RGB}{87,172,255}
\definecolor{thread2}{RGB}{255,201,102}
\definecolor{thread3}{RGB}{255,87,130}

\begin{frame}{Case Study}
	\begin{center}
		\begin{tabular}{llll}
			\begin{tabular}{l}
				Consumer thread \\
				\hline
				\\
				\texttt{LOCK(mx);} \\
				\\
				\texttt{if~(!work\_exists())} \\
				\texttt{~~~~WAIT(cvar,~mx);} \\
				\texttt{work~=~claim\_work();} \\
				\\
				\texttt{UNLOCK(mx);} \\
				\texttt{assert(work~!=~NULL);} \\
			\end{tabular}
			& & &
			\begin{tabular}{l}
				Producer thread \\
				\hline
				\\
				\texttt{LOCK(mx);} \\
				\\
				\texttt{give\_work(work);} \\
				\texttt{SIGNAL(cvar);} \\
				\\
				\texttt{UNLOCK(mx);} \\
				\\
				\\
			\end{tabular}
		\end{tabular}
		\linegap

		\begin{itemize}
			\item See {\bf Paradise Lost} lecture!
			\item {\tt if} vs {\tt while}: Two consumers can race to make one fail.
		\end{itemize}
	\end{center}
\end{frame}
\begin{frame}{Thread Interleavings (``good'' case)}
	\begin{center}
		\begin{tabular}{|l|l|l|}
			\hline
			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
			\hline
			\small \texttt{LOCK(mx);} & & \\
			\small \texttt{if~(!work\_exists())} & & \\
			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\

			& \small \texttt{LOCK(mx);} & \\
			& \small \texttt{give\_work(work);} & \\
			& \small \texttt{SIGNAL(cvar);} & \\
			& \small \texttt{UNLOCK(mx);} & \\

			\small \texttt{work~=~claim\_work();~} & & \\
			\small \texttt{UNLOCK(mx);} & & \\
			\small \texttt{//~got~the~work} {\large \smiley}& & \\

			& & \small \texttt{LOCK(mx);} \\
			& & \small \texttt{if~(!work\_exists())~} \\
			& & \small \texttt{~~~~WAIT(cvar,~mx);} \\
			\hline
		\end{tabular}
	\end{center}
\end{frame}
\begin{frame}{Thread Interleavings (``good'' case, again)}
	\begin{center}
		\begin{tabular}{|l|l|l|}
			\hline
			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
			\hline
			\small \texttt{LOCK(mx);} & & \\
			\small \texttt{if~(!work\_exists())} & & \\
			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\

			& & \small \texttt{LOCK(mx);} \\
			& & \small \texttt{if~(!work\_exists())~} \\
			& & \small \texttt{~~~~WAIT(cvar,~mx);} \\

			& \small \texttt{LOCK(mx);} & \\
			& \small \texttt{give\_work(work);} & \\
			& \small \texttt{SIGNAL(cvar);} & \\
			& \small \texttt{UNLOCK(mx);} & \\

			\small \texttt{work~=~claim\_work();~} & & \\
			\small \texttt{UNLOCK(mx);} & & \\
			\small \texttt{//~got~the~work} {\large \smiley}& & \\
			\hline
		\end{tabular}
	\end{center}
\end{frame}
\begin{frame}{Thread Interleavings (race condition)}
	\begin{center}
		\begin{tabular}{|l|l|l|}
			\hline
			\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
			\hline
			\small \texttt{LOCK(mx);} & & \\
			\small \texttt{if~(!work\_exists())} & & \\
			\small \texttt{~~~~WAIT(cvar,~mx);} & & \\

			& \small \texttt{LOCK(mx);} & \\
			& \small \texttt{give\_work(work);} & \\
			& \small \texttt{SIGNAL(cvar);} & \\
			& \small \texttt{UNLOCK(mx);} & \\

			& & \small \texttt{LOCK(mx);} \\
			& & \small \texttt{work~=~claim\_work();} \\
			& & \small \texttt{UNLOCK(mx);} \\

			\small \texttt{work~=~claim\_work();~} & & \\
			\small \texttt{UNLOCK(mx);} & & \\
			\small \texttt{//~work~was~stolen} {\large \frownie}& & \\
			\hline
		\end{tabular}
	\end{center}
\end{frame}

@@

\begin{frame}{Testing Mechanisms}
	\textbf{Stress testing}: \texttt{largetest}, \texttt{mandelbrot} and friends
	\begin{itemize}
		\item Attempting to exercise as many interleavings as practical
		\item Exposes race conditions at random
		\begin{itemize}
			\item ``If a preemption occurs at just the right time\ldots''
		\end{itemize}
		\item Cryptic panic messages or machine reboots
	\end{itemize}
	\linegap
	What if\ldots
	\begin{itemize}
		\item Make educated guesses about when to preempt
		\item Preempt enough times to run {\em every single} interleaving
		\item Tell the story of what {\em actually happened}.
		\item Overlook fewer bugs!
	\end{itemize}
	% Say: At the end of this lecture, I'm going to be looking for some among you...
\end{frame}

@@

	\linegap
	When do we fear being preempted?
	\begin{itemize}
		\item Threads becoming runnable (\texttt{thr\_create()}, \texttt{cond\_signal()}, etc.)
			\begin{itemize}
				\item Preemptions may cause it to run before we're ready
			\end{itemize}
		\item Synchronization primitives (\texttt{mutex\_lock()}/\texttt{unlock()}, etc.)
			\begin{itemize}
				\item If buggy or used improperly\ldots
			\end{itemize}
		\item Unprotected shared memory accesses (``data races'')% Say: more on this later.
			\begin{itemize}
				\item May result in data structure corruption
			\end{itemize}
	\end{itemize}

@@

	{\bf About Landslide}
	\begin{itemize}
		\item Simics module, which traces:
			\begin{itemize}
				\item Every instruction executed
				\item Every memory access read/written
			\end{itemize}
			% FIXME for next year -- how much slowdown is there? 20x?
		\item Originally supported only P3s; can now test P2s fully-automated
		\item {\em Landslide} shows that your {\em Pebbles} are less stable than you thought.
	\end{itemize}
\end{frame}

@@

%\begin{frame}{Anatomy}
%	\begin{center}
%	\includegraphics[width=0.6\textwidth]{landslide.png}
%	\end{center}
%\end{frame}

\begin{frame}{Identifying Bugs}
	Landslide can {\em definitely discover}:
	\begin{itemize}
		\item Assertion failures % Say: note: the better your asserts are..!
		\item Segfaults
		\item Deadlock
		\item Use-after-free / double-free
	\end{itemize}
	\linegap
	Landslide can {\em reasonably suspect}:
	\begin{itemize}
		\item Infinite loop (halting problem)
		\item Data race bugs
	\end{itemize}
\end{frame}

\begin{frame}{What is a Data Race?}
	A {\bf data race} is a pair of memory accesses between two threads, where:
	\begin{itemize}
		\item At least one of the accesses is a write
		\item The threads are not holding the same mutex
		\item The threads can be reordered (e.g. no \texttt{cond\_signal()} in between)
	\end{itemize}
	\pause
	\linegap

	Data races are {\em not necessarily} bugs, just highly suspicious!
	\begin{itemize}
		\item Is unprotected {\tt foobar\_statistics\_counter++} bad?
		\item What about unprotected {\tt next\_thread\_id++}?
		\item ``If threads interleaved the wrong way here, it {\em might} crash later.''
		\begin{itemize}
			\item Hmmm...
		\end{itemize}
	\end{itemize}
\end{frame}

@@

\begin{frame}{Choosing the Right Preemption Points}
	How can we address exponential state space explosion?
	\pause
	\linegap

@@

\begin{frame}{``Iterative Deepening'' of Preemption Points}
	Goal: Run the best tests for a given CPU budget.
	\linegap

	Based on experience from past 15-410 student volunteers
	\begin{itemize}
		\item Students worked best with an iterative process
		\item ``Start small, then add more preemption points as time allows''
		\item Landslide now automates this process
	\end{itemize}
	\linegap

@@

\breakslide{\Huge Demo}

@@

\section{Landslide for the People}

\begin{frame}{User Study}
	\textbf{Try Landslide on your P2!}
	\begin{itemize}
		\item Bare minimum effort: No more than 1 hour
		\begin{itemize}
			\item Clone a github URL, run setup script, run, answer survey questions
			\item Landslide will automatically report test results
		\end{itemize}
		\item Full study plan: 4-8 hours of active attention
		\begin{itemize}
			\item (Estimated, including time to diagnose and fix bugs)
			\item However, many tests should run passively overnight -- start soon!
		\end{itemize}
	\end{itemize}
	\pause
	\linegap

	Prerequisites
	\begin{itemize}
		\item You {\em must} pass the P2 hurdle before using Landslide.
		\begin{itemize}
			\item {\tt startle}, {\tt agility\_drill}, {\tt cyclone}, {\tt join\_specific\_test}, {\tt thr\_exit\_join}
		\end{itemize}
		\item Must have {\em attempted} several stress tests
		\begin{itemize}
			\item {\tt juggle}, {\tt multitest}, {\tt racer} (for 1 hour), {\tt paraguay}
		\end{itemize}

	\end{itemize}
\end{frame}

\begin{frame}{User Study - Additional Information}
	Human Subjects Research
	\begin{itemize}
		\item CMU IRB has approved this study
		\item Landslide will collect results while you use it
		\begin{itemize}
			\item Record commands issued, take snapshots of your P2 code
			\item All data will be anonymized before publication
		\end{itemize}
		\item No coercion: {\bf \em There is no penalty for not participating.}
		\begin{itemize}
			\item I am not on course staff, cannot influence your grade
			\item Course staff will not have access to study data during semester
		\end{itemize}
	\end{itemize}
	\pause
	\linegap

	Risks \& Benefits
	\begin{itemize}
		\item Benefit: Landslide may help you find/fix bugs, improving your grade!
		\item Risk: Landslide may find no bugs and be a waste of your time.
	\end{itemize}
	% Say: Accordingly, although I hope you will use landslide enough to get good results for me, you should also responsibly allocate time between testing with landslide versus actually implementing your P2.
\end{frame}

% IRB approved, plan on publishing collected data, anonymized
% Expect to spend X time
% Participation has no impact on grade but you will find bugs

\begin{frame}{User Study - How to Participate}
	To participate\dots
	\begin{itemize}
		\item Review this lecture and study information sheet
		\item Meet prerequisites of passing P2 tests
		\item Email me ({\tt bblum@cs.cmu.edu}) for instructions, survey
		\item Use Landslide in addition to stress tests until P2 is due!
	\end{itemize}
	\linegap

	Watch your email for more details!
\end{frame}

\section{Bonus Slides}

\begin{frame}{Coping with State Space Explosion}
	Serious problem: State spaces grow exponentially
	\begin{itemize}
		\item With $p$ preemption points and $k$ runnable threads, size $p^k$.
		\item Threatens our ability to explore everything.
		\item Fortunately, some sequences result in identical states.
	\end{itemize}
	\linegap

\subsection{Experimental Results}

@@

\begin{frame}{More Preliminary Results}
	\textbf{Is dynamically adding ``data race'' PPs effective?}
	\begin{itemize}
		\item 3 among reported bugs (9-15\%) required data races to expose.
	\end{itemize}
	\linegap
